\documentclass[phd,tbottom,tnosig]{Athesis}

\usepackage{mystyle}

\usepackage{multirow}
\usepackage{afterpage}
\usepackage{hhline}
\usepackage{url}
%\usepackage{hyperref} % see the file hyperref.cfg for options
%\usepackage{footmisc} % provides \footref and \mpfootnotemark commands
\usepackage{xspace}
\usepackage[numbers,sort&compress]{natbib}

% Latin abbreviations
\providecommand*{\eg}{\emph{e.\,g.}\xspace}%
\providecommand*{\ie}{\emph{i.\,e.}\xspace}%

\hyphenation{put words here which LaTeX does not hy-phen-ate pro-per-ly}

\author{James Zuber}%
\title{Probing the Knowable Unknown}%

\program{Applied Mathematics and Statistics}%

\director{Joseph S. B. Mitchell}{Distinguished Professor, Department of Applied Mathematics and Statistics}%
\chairman{Esther Arkin}{Professor, Department of Applied Mathematics and Statistics}%
\fstmember{Steven Skiena}{Distinguished Teaching Professor, Department of Computer Science}%
\sndmember{Jie Gao}{Associate Professor, Department of Computer Science}%
\dean{Charles Taber}%

\begin{document}

\singlespacing %
\pagenumbering{roman} %

%\maketitle %
\makeapproval %

\begin{abstract}
\input{abstract}
\end{abstract}

\begin{dedication}
  To JC.
\end{dedication}

\tableofcontents %
%\listoffigures %
%\listoftables %

\pagestyle{thesis}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\pagenumbering{arabic}

\chapter{Airplane Seating}{

\begin{centering}
\par{\large \bf Summary:}

\end{centering}

\par{We imagine passengers on an airplane being asked to swap seats so that a couple or family can sit together. This paper focuses on minimizing the number of passenger exchanges to make all families on a plane happy. We give a hardness proof for the general case of unbounded, heterogeneous family sizes and constant factor approximation heuristics for smaller families on idealized planes.} 
}
\input{airplaneBody}

\chapter{Genomic Interval Algorithm Analysis}{


\begin{centering}
\par{\large \bf Summary:}

\end{centering}

\par{Using a specific mathematical model, we treat DNA copy number analysis as an optimization problem. We devise a polynomaial time approximation scheme, two greedy heuristics and a binary programming model to solve the model exactly. These algorithms are analyzed for correctness and then their performance is tested using synthetic data.} 
}
\input{intervalsBody}

\chapter{The Waiter Problem in One Dimension}{
\begin{centering}
\par{\it James Zuber}

\par{\large \bf Summary:}

\end{centering}

\par{The waiter problem is a discrete packing problem where the objective is to contain the partial centers of mass of objects placed on a line within a small interval. We introduce the problem, prove its NP completeness, and present both a bound on performance and a 2.7 factor approximation heuristic to solve the problem.}
}

\input{waiter1DBody}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliographystyle{unsrtnat}
\bibliography{airplanePapers,intervalPapers,waiterPapers}

\end{document}
